---
title: "Working with the geneset management system GeMS"
author: "Jitao David Zhang"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  rmarkdown::html_vignette:
    fig_caption: yes
  html_document:
    df_print: paged
    theme: spacelab
    mathjax: default
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: inline
params:
  echo: yes
  relative: FALSE
vignette: >
  %\VignetteIndexEntry{Working with the geneset management system GeMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(ribiosGSEA)
library(BioQC)
doEval <- ribiosGSEA::isGeMSReachable()
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE,
                      eval = doEval,
                      fig.height=6, fig.width=6)
```

# Background

The Geneset Managment System ([GeMS, link visible within Roche](https://rochewiki.roche.com/confluence/display/BEDA/Gene+sets+and+signatures+-+GeMS)), as its name suggests, is a system to manage genesets.

This document shows basic operations provided to work with the sytsem from R console. 

Note that the evaluation only takes place when GeMS is reachable, because otherwise the compilation of this vignette  will fail when the webservice is not available. 

# Examples

## Task 1: show genesets provided by a user

The following command fetches genesets associated with a given user, *test* in this case. If not parameter is given, the genesets of the current user is provided.

```{r getGeneSets, eval=FALSE}
getUserSetsFromGeMS(user="test")
```

## Task 2: insert new genesets to GeMS

New genesets can be inserted into GeMS, using the *GmtList* data structure defined in the *BioQC* package, which is essentially a list of lists, which in turn contains three elements: *name*, *desc*, and *genes*. See example below.

```{r insertGeneSets, eval=FALSE}
testGmt <- BioQC::GmtList(list(
  list(name="Test gene set 1", 
       desc="Test", genes=c("CXCL13", "TNFRSF1B", "RGS2")),
  list(name="Test gene set 2", desc="Test", 
       genes=c("ACADSB", "AFTPH", "ARL1"))
))
insertGmtListToGeMS(testGmt, geneFormat=0, source="Test", xref="PMID:30397336", user="test")
```

Now we check the genesets of the user *test*, we expect to see the two gene sets that we inserted.

```{r getGeneSetsAfterInsert, eval=FALSE}
getUserSetsFromGeMS(user="test")
```

Alternative to manual specificiation, *GmtList* can be constructed by reading a valid GMT file with the function *readGmt* in *BioQC*.

## Task 3: Remove a geneset

The following example shows how a geneset can be removed
```{r removeGeneSet, eval=FALSE}
removeFromGeMS(setName=c("Test gene set 1", "Test gene set 2"),
               source="Test", user="test")
```

Now we check the genesets of the user *test*, we expect to see the two gene sets that we previously inserted should be gone.

```{r getGeneSetsAfterRemove eval=FALSE}
getUserSetsFromGeMS(user="test")
```

# Conclusions

This is a very brief introduction of how to working with GeMS within R. In case of questions or suggestions, please reach out to [Jitao David Zhang](mailto:jitao_david.zhang@roche.com) or [Laura Badi](mailto:laura.badi@roche.com).


# Session information

```{r}
sessionInfo()
```






